---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---

## Install and Load Libraries
```{r message=FALSE}
# Auto Install Packages
list.of.packages <- c("dplyr", "ggplot2", "knitr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)

## Bring in the libs.
library(paste0(list.of.packages, collapse = ', '))
```

## Loading and preprocessing the data
```{r}
    # Load the data from CSV
    activity <- read.csv((unz("activity.zip", "activity.csv")))
    
    # Turn the date column into actual dates.    
    activity <- activity %>%
        mutate(date = as.Date(date))
```



## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r sumsteps, fig.height = 4, fig.width = 6}
    daily_activity <- activity %>%
        filter(!is.na(steps)) %>%
        group_by(date) %>%
        summarise(sumSteps = sum(steps, na.rm = TRUE))

    g <- ggplot(data = daily_activity, aes(sumSteps)) + 
        labs(title="Total Number of Steps Taken Per Day, Histogram") + 
        geom_histogram()

    print(g)
## Todo - come back for axis labels.
```

#### Mean Steps Taken Per Day: 
```{r results='asis'}
    meanSteps <- mean(daily_activity$sumSteps, na.rm = TRUE)
    # Make it a bit easier to read than the super long number.
    round(meanSteps, 3)
```

##### Median Steps Per Day
``` {r results='asis'}
    medianSteps <- median(daily_activity$sumSteps, na.rm = TRUE)
    medianSteps
```




## What is the average daily activity pattern?

##### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r activity_pattern, fig.height = 4, fig.width = 6}
    ## Once again ignoring NAs
    clean_activity <- filter(activity, !is.na(steps))
    
    intervalCount <- count(clean_activity, interval)
   
    activity_pattern <- activity %>%
        group_by(interval) %>%
        summarise(sum_steps = sum(steps, na.rm = TRUE)) %>%
        inner_join(intervalCount, c("interval")) %>%
        mutate(avg_steps_per_interval = sum_steps / n)

    ## Using ggplot2, since assignment allows using any plotting system.
    g <- ggplot(
            data = activity_pattern, 
            aes(x = interval, y = avg_steps_per_interval)
        ) + 
        labs(
            title = "Average Daily Activity Pattern",
            x = "Interval",
            y = "Average Steps Per Interval"
        ) + 
        geom_line()
    print(g)
## TODO -- come back for x axis labels for 5 minute intervals.

#plot(activity_pattern$interval, activity_pattern$avg_steps_per_interval, type="l")

```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
    ap_ordered <- arrange(activity_pattern, desc(avg_steps_per_interval))
    ap_high_interval <- ap_ordered[1, 1]
    ap_high_interval_steps <- ap_ordered[1, 4]
```

The 5 minute interval `r ap_high_interval` has the highest number of average steps per day, with `r round(ap_high_interval_steps, digits=2)` steps per interval. 

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#####1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
total_bad <- activity[!complete.cases(activity), ]
nrow(total_bad)
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Are there differences in activity patterns between weekdays and weekends?
